### 2022-10-13 AISTATS2023/paper.tex
latest

Similarly to~\citep{aas2021explaining, herbinger2022repid} we first
use simulation examples for evaluation. The first group
(Section~\ref{sec:simulation-examples-1}) compares UALE against
PDP-ICE in modeling the average effect and the uncertainty. The second
group (Section~\ref{sec:simulation-examples-2}) compares UALE
approximation with optimal bin-splitting against the fixed-size
alternative.

@@

The aim of this example is to highlight that when features are
correlated, PDP and ICE fail in quantifying the average effect and the
heterogeneity, even when the black-box function is relatatively
simple.

We use the following distribution:
\(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\) and the following piecewise
linear function:

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $f_1(\mathbf{x}) < \frac{1}{2}$ }\\
                    \frac{1}{2} - f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $\frac{1}{2} \leq f_2(\mathbf{x}) < 1$}\\
                    \alpha f_1(\mathbf{x}) &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_1(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_2(\mathbf{x}) = x_1x_3\). The linear term \(f_1\) includes the two
correlated features and the interaction term \(f_2\) the two
non-correlated.  We evaluate the effect computed by UALE and PDP-ICE
in three cases; (a) without interaction (\(\alpha=0\)) and equal
weights (\(a_1=a_2\)), (b) without interaction (\(\alpha=0\)) and
different weigths (\( a_1 \neq a_2 \)) and (c) with interaction
(\(\alpha \neq 0\)) with equal weights (\(a_1=a_2\)).

In all cases, we assume that because the \todo{give ground truth
  effect}. We compare it against the approximations provided by each
method, after generating \(N=500\) samples.

@@

In this set-up, \(\alpha=0\) (no interaction) and the \(a_1=a_2=1\)
(equal weights). The ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\)
is: \(x_1\) in \([0, 0.25]\), \(-x_1\) in \([0.25, 0.5]\) and \(0\) in
\([0.5, 1]\). The uncertainty is
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) in \([0,1]\) because \(x_1\)
does not interact with ohter variables. In
Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP effect is
wrong and ICE plots show heterogeneous effects. In contrast, UALE
models correctly both the average effect and the
uncertainty. Furthermore, we observe that bin-splitting is accurate
(zero cumulative error \(\mathcal{E}_{\mathcal{Z}}\)).

>>>>

We perform a two-fold evaluation of UALE. The first group
(Section~\ref{sec:simulation-examples-1}) compares UALE with PDP-ICE,
the main feature effect method that also quantifies the heterogeneity
of instance-based effects. The second group
(Section~\ref{sec:simulation-examples-2}) evaluates UALE
approximation, i.e., compares UALE's optimal bin-splitting against the
fixed-size alternative.

@@

This a qualitative example with the aim to highlight that PDP-ICE is
vulnerable to misleading estimations of both the average effect and
the uncertainty, even when the black-box function is relatatively
simple.

We use the data generating distribution
\(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\) and the following piecewise
linear function

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $f_1(\mathbf{x}) < \frac{1}{2}$ }\\
                    \frac{1}{2} - f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $\frac{1}{2} \leq f_1(\mathbf{x}) < 1$}\\
                    \alpha f_2(\mathbf{x}) &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_1(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_2(\mathbf{x}) = x_1x_3\). The part \(f_1\) is a linear function of
the two correlated features, \(x_1, x_2\), and \(f_2\) is an
interaction term between the two non-correlated, \(x_1, x_3\).  We
evaluate the effect computed by UALE and PDP-ICE in three cases; (a)
without interaction (\(\alpha=0\)) and equal weights (\(a_1=a_2\)),
(b) without interaction (\(\alpha=0\)) and different weigths
(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha > 0\)) with
equal weights (\(a_1=a_2\)).

In the general case, there is no unique ground-truth feature effect
and uncertainty and each method defines the effect in a different
way. We select this simple case, to determine the ground-truth,
exploiting two facts. First, in \(p(\xb)\) it holds that
\(x_1=x_2\). Therefore, knowing the value of either \(x_1\) or
\(x_2\), we know the closed-form of \(f(\xb)\), too. For example, if
\(a_1=a_2=1\) and \(x_1 = \frac{1}{6}\), then
\(f_1(\xb) = \frac{1}{3}\), so
\(f(\xb) = f_1(\xb) + \alpha f_2(\xb) \). Afterwards, we use the
properties of interpretable models~\citep{molnar2022} to determine the
ground-truth. In the example above, if also \(a=0\), then
\(f(\xb) = x_1 + x_2\) which is an interpretable linear model. For the
uncertainty, we use the fact that under no-interactions, \(a=0\), the
uncertainty must be zero (no heteregoneity). We discuss separately the
case with \(a>0\). Finally, we want to also show that UALE's
approximation is accurate, due to appropriate bin splitting.

@@

In this set-up, \(\alpha=0\) (no interaction) and the \(a_1=a_2=1\)
(equal weights). We examine only the effect of feature \(x_1\),
because it is exactly the same with \(x_2\). Based on the discussion
above, the ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is:
\(x_1\) in \([0, \frac{1}{4}]\), \(-x_1\) in
\([\frac{1}{4}, \frac{1}{2}]\) and \(0\) in \([\frac{1}{2},
1]\). Because \(x_1\) does not interact with any other variable, the
uncertainty is \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) in
\([0,1]\). In Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP
effect is wrong and ICE plots show heterogeneous effects. In contrast,
UALE models correctly both the average effect and the absence of
uncertainty. Furthermore, we observe that bin-splitting is accurate
(zero cumulative error \(\mathcal{E}_{\mathcal{Z}}\)).
